Software architectures to integrate workflow engines in science gateways

### 2016-05-14 paper.tex
Up to Evaluation

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years, as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

  the infrastructure, which may require administrative privileges;

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

>>>>

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed \note{vladimir}{is this a good word for this context, 'informed'? maybe 'based on'?} by our experience in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years, as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures \note{vladimir}{why is it exactly six? are these architectures the ones that happened to be implemented during various projects, or is it a full set of all possibilities (I believe not)? I think, it must be stated clearly why the focus is given to these particular ones - we are reviewing and evaluating the approaches that have been used already in our projects} within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features \note{vladimir}{can we explain why exactly these metrics are used but not anything else? I think, we need to justify that these ones are especially important for the SG case; it would be nice to refer to some related research on software architecture metrics, Google gives about 2 million pages on this, where you can find metrics like cohesion, depth of scenario and so on}. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
and WS-PGRADE~\cite{Kacsuk2012} \note{vladimir}{As a past developer of it, I would suggest to refer to WS-VLAM as well, despite it is not very famous, if you do not mind :)}. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

  the infrastructure, which may require administrative privileges \note{vladimir}{or bundled in packages considering all needed dependencies than can be deployed on the fly during the workflow execution};

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. \note{vladimir}{Are we talking only about importing workflows from other wf languages? I would suggest to modify the title of the section to 'Workflow conversion and import' then, as some gateways use the term 'import' to import their own native workflows, like WS-PGRADE, and using import only in the context of converting workflows can be confusing} Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
\note{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered. One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly. So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.
